## GDAA 1001 -- Fundamentals of Geospatial Data Analytics

## Project #2

## 
Title:       Exploratory Data Analysis (EDA)

## Author:    Christian Akpanya

## Date: November  8th 2024

## 
Introduction

#### This document presents an exploratory data analysis (EDA) of the selected data set.  The data set was originally pre processed  in an ETL  project. The analysis aims to uncover patterns, trends, and relationships in the data to guide further modeling or decision-making processes.

## Load Libraries and Data

```{r setup, include=TRUE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)

# Load the dataset
data <- read.csv("C:\\GDAA\\Project2\\data\\processed_data\\df_clean.csv")

# Display the structure of the dataset
str(data)
summary(data)
```

1.  <div>

    ## Single Numeric Variable Analysis

    </div>

    ### Descriptive Statistics

```{r}
# Select numeric columns for analysis
numeric_vars <- select_if(data, is.numeric)

# Compute descriptive statistics
stats <- numeric_vars %>%
  summarise_all(list(
    mean = mean,
    median = median,
    sd = sd,
    IQR = IQR,
    min = min,
    max = max
  ))

# Print the computed statistics
print(stats)


```

## Visualizations

#### Density Plot

```{r}
ggplot(data, aes(x = numeric_vars[[1]])) +
  geom_density(fill = "blue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Density Plot", x = "Variable", y = "Density")

```

### Histogram

```{r}
ggplot(data, aes(x = numeric_vars[[1]])) +
  geom_histogram(binwidth = 5, fill = "orange", color = "black") +
  theme_minimal() +
  labs(title = "Histogram", x = "Variable", y = "Frequency")

```

### Box Plot

```{r}
ggplot(data, aes(y = numeric_vars[[1]])) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Box Plot", y = "Variable")

```

###  Correlation Analysis

```{r}
# Compute correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")
print(cor_matrix)

# Scatterplot matrix
ggpairs(numeric_vars)

```

## Single Categorical Variable Analysis

### Frequency and Proportion Counts

```{r}
# Select categorical columns for analysis
categorical_vars <- select_if(data, is.factor)

freq_table <- categorical_vars %>%
  summarise_all(list(freq = ~length(.)))

print(freq_table)


```

### Bar Plot

```{r}
categorical_vars <- select_if(data, is.factor)

# Ensure there are categorical variables before plotting
if (length(categorical_vars) > 0) {
  ggplot(data, aes(x = categorical_vars[[1]])) +
    geom_bar(fill = "green", alpha = 0.7) +
    theme_minimal() +
    labs(title = "Bar Plot", x = "Category", y = "Frequency")
} else {
  print("No categorical variables found in the dataset.")
}
data$facility_type <- as.factor(data$facility_type)


ggplot(data, aes(x = facility_type)) +
  geom_bar(fill = "green", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Bar Plot of Facility Types", x = "Facility Type", y = "Frequency")





```

##  Multiple Categorical Variable Analysis

### Heatmap of Associations

```{r}
#Bin Latitude into Regions

data$latitude_bin <- cut(
  data$latitude,
  breaks = 4,
  labels = c("North", "Central North", "Central South", "South")
)

# Create a contingency table
contingency_table <- table(data$facility_type, data$latitude_bin)
print(contingency_table)


# Heatmap visualization
ggplot(as.data.frame(contingency_table), aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(title = "Heatmap of Associations", x = "Variable 1", y = "Variable 2", fill = "Frequency")

```

### Frequency and Proportion Counts

```{r}
# Frequency and Proportion Counts for 'facility_type'
facility_type_counts <- data %>%
  count(facility_type) %>%
  mutate(proportion = n / sum(n))

print(facility_type_counts)

```

### Simple Bar Plot

```{r}
ggplot(data, aes(x = facility_type)) +
  geom_bar(fill = "skyblue", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Frequency of Facility Types", x = "Facility Type", y = "Count")

```

### Enhanced Bar Plots

Overlay additional information by stacking bars for categories.

used `latitude_bin` (created earlier) to show facility distribution across regions.

```{r}
# Create a binned version of latitude
data$latitude_bin <- cut(
  data$latitude,
  breaks = 4, # Number of bins (e.g., 4 regions)
  labels = c("North", "Central North", "Central South", "South")
)

# Verify the new column
print(table(data$latitude_bin))

# Stacked Bar Plot
ggplot(data, aes(x = facility_type, fill = latitude_bin)) +
  geom_bar(position = "stack", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Stacked Bar Plot of Facility Types by Latitude Bin",
       x = "Facility Type", y = "Count", fill = "Latitude Bin")

```

#### **Dodged Bar Plot**:

Display the categories side by side for easier comparison.

```{r}
# Dodged Bar Plot
ggplot(data, aes(x = facility_type, fill = latitude_bin)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Dodged Bar Plot of Facility Types by Latitude Bin",
       x = "Facility Type", y = "Count", fill = "Latitude Bin")


```

#### **Filled Bar Plot**:

Show proportions instead of raw counts using the `position = "fill"` argument.

```{r}
# Filled Bar Plot
ggplot(data, aes(x = facility_type, fill = latitude_bin)) +
  geom_bar(position = "fill", alpha = 0.8) +
  theme_minimal() +
  labs(title = "Filled Bar Plot of Facility Types by Latitude Bin",
       x = "Facility Type", y = "Proportion", fill = "Latitude Bin")

```

### Interpretation:

-   The **simple bar plot** shows the overall distribution of facility types.

-   The **stacked bar plot** helps visualize the distribution across latitude bins.

-   The **dodged bar plot** makes it easier to compare categories side by side.

-   The **filled bar plot** highlights proportions, providing insight into relative distributions.

### **Correspondence Analysis**

Perform correspondence analysis if deeper exploration is required.

```{r}
library(FactoMineR)
library(factoextra)

# Perform correspondence analysis
correspondence_result <- CA(contingency_table, graph = FALSE)

# Visualize correspondence analysis
fviz_ca_biplot(correspondence_result, repel = TRUE) +
  labs(title = "Correspondence Analysis of Facility Types and Latitude Bins")


```

## Conclusion

This EDA has identified key trends and relationships in the data-set. Insights from this analysis will guide subsequent modeling or decision-making processes.
